Convolutional equivalent layer for gravity data processing

### 2020-04-30 manuscript/abstract.tex
include FFT results in conclusion

We have developed an efficient and very fast equivalent-layer technique for gravity data processing by modifying an 
iterative method grounded on excess mass constraint that does not require the solution of linear systems. 
Taking advantage of the symmetric Block-Toeplitz Toeplitz-block (BTTB) structure of the sensitivity matrix, that raises 
when regular grids of observation points and equivalent sources (point masses) are used to set up a fictitious 
equivalent layer, we have developed an algorithm which greatly reduces the number of flops and RAM memory necessary 
to estimate a 2D mass distribution over the equivalent layer. The structure of symmetric BTTB matrix consists of 
the elements of the first column of the sensitivity matrix, which in turn can be embedded into a symmetric 
Block-Circulant Circulant-Block (BCCB) matrix. Likewise, only the first column of the BCCB matrix is needed 
to reconstruct the full sensitivity matrix completely. From the first column of BCCB matrix, its eigenvalues 
can be calculated using the 2D Fast Fourier Transform (2D FFT), which can be used to readily 
compute the matrix-vector product of the forward modeling in the fast equivalent-layer technique. As a result, 
our method is efficient to process very large datasets using either fine- or mid-grid meshes. The larger the 
dataset, the faster and more efficient our method becomes compared to the available equivalent-layer techniques. 
Tests with synthetic data demonstrate the ability of our method to satisfactorily upward-
and downward-continuing the gravity data.
Our results show very small border effects and noise amplification compared to those 
produced by the classical approach in the Fourier domain.
Besides, they show that while the running time of our method is $\approx 30.9$ seconds for 
processing $N = 1\,000\,000$ observations, the fast equivalent-layer technique spent 
$\approx 46.8$ seconds with $N = 22\,500$.
A test with field data from Caraj{\'a}s province, Brazil, illustrates the low computational 
cost of our method to process a large data set composed of $N = 250\,000$ observations.

>>>>

We have developed an efficient and very fast equivalent-layer technique for gravity data processing by modifying an 
iterative method grounded on excess mass constraint that does not require the solution of linear systems. 
Taking advantage of the symmetric Block-Toeplitz Toeplitz-block (BTTB) structure of the sensitivity matrix, that raises 
when regular grids of observation points and equivalent sources (point masses) are used to set up a fictitious 
equivalent layer, we have developed an algorithm which greatly reduces the number of flops and RAM memory necessary 
to estimate a 2D mass distribution over the equivalent layer. The structure of symmetric BTTB matrix consists of 
the elements of the first column of the sensitivity matrix, which in turn can be embedded into a symmetric 
Block-Circulant Circulant-Block (BCCB) matrix. Likewise, only the first column of the BCCB matrix is needed 
to reconstruct the full sensitivity matrix completely. From the first column of BCCB matrix, its eigenvalues 
can be calculated using the 2D Fast Fourier Transform (2D FFT), which can be used to readily 
compute the matrix-vector product of the forward modeling in the fast equivalent-layer technique. As a result, 
our method is efficient to process very large datasets using either fine- or mid-grid meshes. The larger the 
dataset, the faster and more efficient our method becomes compared to the available equivalent-layer techniques. 
Tests with synthetic data demonstrate the ability of our method to satisfactorily upward-
and downward-continuing the gravity data.
Our results show very small border effects and noise amplification compared to those 
produced by the classical approach in the Fourier domain.
Besides, they show that while the running time of our method is $\approx 30.9$ seconds for 
processing $N = 1\,000\,000$ observations, the fast equivalent-layer technique spent 
$\approx 46.8$ seconds with $N = 22\,500$.
A test with field data from Caraj{\'a}s Province, Brazil, illustrates the low computational 
cost of our method to process a large data set composed of $N = 250\,000$ observations.

### 2020-04-30 manuscript/conclusions.tex
include FFT results in conclusion

We show that the sensitivity matrix associated with the equivalent layer technique 
for processing gravity data has a BTTB structure when the following conditions are
satisfied: (i) the observed data are disposed at a regular grid on a horizontal plane
and (ii) the equivalent sources (point masses) are placed at a constant depth, one directly 
beneath each observation point.
By exploring the BTTB structure of the sensitivity matrix, we formulate the forward modeling as a 
fast FFT convolution that requires only one column of the sensitivity matrix 
and propose an efficient approach for optimizing the computational time of an 
iterative method called fast equivalent-layer technique.
This iterative method is grounded on the excess mass constraint and does not demand the solution of 
linear systems.
Our approach greatly reduces the number of flops and the RAM memory necessary to estimate a 
2D mass distribution within a planar equivalent layer that fits the observed gravity data. 
For example, the number of flops of the fast equivalent-layer technique is reduced by two orders 
of magnitude when processing one million of observations by using our approach. 
Traditionally, such amount of data impractically requires $7.6$ Terabytes of RAM memory to handle 
the full sensitivity matrix whereas our method takes only $61.035$ Megabytes.
This drastic reduction comes from the fact that, by using only one column of the sensitivity matrix,
our method is able to compute the gravity data produced by the whole equivalent layer with a single 
point mass via FFT.
We have successfully applied the proposed method to compute the upward/downward continuation of 
synthetic gravity data. Application to field data over the Caraj{\'a}s Province, north of Brazil, 
confirms the potential of our approach in processing a gravity data set with $250\,000$ observations 
in approximately $0.2$ seconds. Further studies need to be carried out in order to apply our 
method to process other potential-field data.

>>>>

We show that the sensitivity matrix associated with the equivalent layer technique 
for processing gravity data has a BTTB structure when the following conditions are
satisfied: (i) the observed data are disposed at a regular grid on a horizontal plane
and (ii) the equivalent sources (point masses) are placed at a constant depth, one directly 
beneath each observation point.
By exploring the BTTB structure of the sensitivity matrix, we formulate the forward modeling as a 
fast FFT convolution that requires only one column of the sensitivity matrix 
and propose an efficient approach for optimizing the computational time of an 
iterative method called fast equivalent-layer technique.
This iterative method is grounded on the excess mass constraint and does not demand the solution of 
linear systems.
Compared to the fast equivalent-layer technique, 
our approach greatly reduces the number of flops and the RAM memory necessary to estimate a 
2D mass distribution within a planar equivalent layer that fits the observed gravity data. 
For example, the number of flops is reduced by two orders 
of magnitude when processing one million of observations by using our approach. 
Traditionally, such amount of data impractically requires $7.6$ Terabytes of RAM memory to handle 
the full sensitivity matrix whereas our method takes only $61.035$ Megabytes.
This drastic reduction comes from the fact that, by using only one column of the sensitivity matrix,
our method is able to compute the gravity data produced by the whole equivalent layer with a single 
point mass via FFT.
We have successfully applied the proposed method to compute the upward/downward continuation of 
synthetic gravity data. 
Our results show that our method produces considerably smaller border effects and 
noise amplification compared to the classical Fourier approach, 
which computes the Fourier transform of the observed data.
Application to field data over the Caraj{\'a}s Province, north of Brazil, 
confirms the potential of our approach in processing a gravity data set with $250\,000$ observations in approximately $0.2$ seconds. Further studies need to be carried out in order to apply our method to process other potential-field data.
